An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-23 EconGrowth.Rnw
Added some to-do items.
Now including code to print CES parameter tables in the appendix. Not quite working yet, unfortunately. The values of the sigma parameters are messed up.

\section*{To-do List}
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item Heun: Add CES production function fits
      \begin{itemize}
      \item \st{GDP comparison graph}
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Deal with missing covariance numbers when columns have unequal length.
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email on 4 Feb 2013 but have not heard a response.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the LINEX model
      \begin{itemize}
      \item Zambia, upper bound on $c_t$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item Heun: x-axis labels only on bottom for 9-panel lattice graphs
\item Heun: adjust y axis limits on $a_0$ and $c_t$ parameter graphs for the LINEX model AFTER we finish debugging the LINEX models.
\item de Wit: provide funding acknowledgement, if necessary.
\item de Wit: review, critique, and contextualize.
\item \st{Heun: Ask Martin if he can supply an ``Economics'' affiliation. Will probably make a stronger paper.}
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item \st{Heun: Move tables to an appendix at the end.}
\item \st{Heun: Add acknowledgements (Dad, Pruim)}
\item \st{Heun: Add $u$ parameter table and graph}
\item \st{Heun: Add covariance metrics.}
\item \st{Heun: Add single factor fits}
      \begin{itemize}
      \item \st{Correct sizes for parameter graphs for each country}
      \item \st{Parameter graphs ($\lambda$ and $m$) for each country}
      \item \st{Parameter tables for each country in the appendix}
      \item \st{GDP comparison graph}
      \item \st{AIC rows in the AIC table}
      \end{itemize}
\item \st{Heun: Add setup variables for figure height and width where not already included.}
\item \st{Heun: Add LINEX production function fits}
      \begin{itemize}
      \item \st{Parameter graphs for each country}
      \item \st{Parameter tables for each country in the appendix}
      \item \st{AIC rows in the AIC table}
      \item \st{GDP comparison graph}
      \end{itemize}
\end{itemize}

@@

cesData <- function(countryName, energyType){
  energyColumnName <- paste("i", energyType, sep="")
  # Load the data that we need.
  dataTable <- loadData(countryName)

  # Establish guess values for phi beta, zeta, lambda_L and lambda_E.
  phiGuess <- -20
  betaGuess <- 0.5 # a typical value for beta (exponent on labor)
  zetaGuess <- 0.0004 # a small value
  lambda_LGuess <- 0.007 #assuming no technical progress on the labor-capital portion of the function
  lambda_EGuess <- 0.008 #assuming no technical progress on the energy portion of the function

  # Runs a non-linear least squares fit to the data with constraints
  modelCES <- nls(iGDP ~ ((1-zeta) * (exp(lambda_L*iYear) * iCapStk^(1-beta) * iLabor^beta)^phi 
                           + zeta*(exp(lambda_E*iYear) * iQ)^phi)^(1/phi), 
                   algorithm = "port",
                   control = nls.control(maxiter = 500, tol = 1e-06, minFactor = 1/1024, 
                                         printEval = FALSE, warnOnly = FALSE),
                   start = list(phi=phiGuess, beta=betaGuess, zeta=zetaGuess, lambda_L=lambda_LGuess, 
                                lambda_E=lambda_EGuess),
                   lower = list(phi=-Inf, beta=0, zeta=0, lambda_L=-Inf, lambda_E=-Inf),
                   upper = list(phi=0, beta=1, zeta=1, lambda_L=Inf, lambda_E=Inf),
                   data=dataTable)

  aicCES <- AIC(modelCES, k=2) # Checks validity of the model. AIC stands for Akaike's Information Criterion
  print(aicCES)

  # Gives the nls summary table
  summaryCES <- summary(modelCES) # Gives the nls summary table
  print(summaryCES)

  # Provides confidence intervals on phi, beta, zeta, lambda_L, and lambda_E. But, we need the CI on alpha.
  ciCES <- confint(modelCES, level = ciLevel)
  print(ciCES)

  # Get the estimate for alpha
  beta <- as.numeric(coef(modelCES)["beta"])
  alpha <- 1.0 - beta
  alpha.est <- deltaMethod(modelCES, "1 - beta") # Estimates alpha and its standard error (SE).
  print(alpha.est)

  # Now calculate a confidence interval on alpha
  dofCES <- summaryCES$df[2]
  print(dofCES) # Gives the degrees of freedom for the model.
  tvalCES <- qt(ciHalfLevel, df = dofCES); tvalCES
  # Get confidence intervals for each parameter in the model
  alphaCICES <- with(alpha.est, Estimate + c(-1.0, 1.0) * tvalCES * SE) # CI on alpha.
  print(alphaCICES)

  # Assemble the data into data frames for the table.
  estCES <- data.frame(phi = coef(modelCES)["phi"], alpha = alpha, 
                       beta = coef(modelCES)["beta"], zeta = coef(modelCES)["zeta"], 
                       lambda_L = coef(modelCES)["lambda_L"], lambda_E = coef(modelCES)["lambda_E"])
  row.names(estCES) <- paste("CES with ", energyType, sep="")
  #print(estCES)  
  # The [1] subscripts pick off the lower confidence interval
  lowerCES <- data.frame(phi = ciCES["phi","2.5%"], alpha = alphaCICES[1], 
                         beta = ciCES["beta", "2.5%"], zeta = ciCES["zeta", "2.5%"],
                         lambda_L = ciCES["lambda_L", "2.5%"], lambda_E = ciCES["lambda_E", "2.5%"])
  row.names(lowerCES) <- "- 95% CI"
  # The [2] subscripts pick off the lower confidence interval
  upperCES <- data.frame(phi = ciCES["phi","97.5%"], alpha = alphaCICES[2], 
                         beta = ciCES["beta", "97.5%"], zeta = ciCES["zeta", "97.5%"],
                         lambda_L = ciCES["lambda_L", "97.5%"], lambda_E = ciCES["lambda_E", "97.5%"])
  row.names(upperCES) <- "+ 95% CI"

  # Now create the data for a table.
  dataCES <- rbind(upperCES, estCES, lowerCES)
  return(dataCES)
}

####################################
# Creates a LaTeX printable table from the CES data. This function first calls cesData.
#
# countryName is a string containint the 2-letter abbreviation for the country, e.g. "US" or "CN"
# energyType is a string to be used in table captions reprsenting the type of energy. Typically, "Q", "X", or "U"
#
# returns a printable LaTeX table from xtable.
##
cesTable <- function(countryName, energyType){
  dataCESe <- cesData(countryName, energyType)
  tableCESq <- xtable(dataCESe, caption=paste(countryName, ", 1980-2011.", sep=""), digit = c(4, 1, 2, 2, 6, 5, 5))
}
@

@@

>>>>

\section*{To-do List}
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item Heun: Add CES production function fits
      \begin{itemize}
      \item \st{GDP comparison graph}
      \item Convert $\rho$ variables in the \citet{Henningsen:2011td} CES production function to $\sigma$ variables.
      \item parameter graphs for each country
      \item parameter tables for each country in the appendix
      \item AIC rows in the AIC table
      \end{itemize}
\item Heun: Deal with missing covariance numbers when columns have unequal length. Use the \texttt{subset} command.
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email on 4 Feb 2013 but have not heard a response.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the LINEX model
      \begin{itemize}
      \item Zambia, upper bound on $c_t$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item Heun: x-axis labels only on bottom for 9-panel lattice graphs
\item Heun: adjust y axis limits on $a_0$ and $c_t$ parameter graphs for the LINEX model AFTER we finish debugging the LINEX models.
\item de Wit: provide funding acknowledgement, if necessary.
\item de Wit: review, critique, and contextualize.
\item \st{Heun: Ask Martin if he can supply an ``Economics'' affiliation. Will probably make a stronger paper.}
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item \st{Heun: Move tables to an appendix at the end.}
\item \st{Heun: Add acknowledgements (Dad, Pruim)}
\item \st{Heun: Add $u$ parameter table and graph}
\item \st{Heun: Add covariance metrics.}
\item \st{Heun: Add single factor fits}
      \begin{itemize}
      \item \st{Correct sizes for parameter graphs for each country}
      \item \st{Parameter graphs ($\lambda$ and $m$) for each country}
      \item \st{Parameter tables for each country in the appendix}
      \item \st{GDP comparison graph}
      \item \st{AIC rows in the AIC table}
      \end{itemize}
\item \st{Heun: Add setup variables for figure height and width where not already included.}
\item \st{Heun: Add LINEX production function fits}
      \begin{itemize}
      \item \st{Parameter graphs for each country}
      \item \st{Parameter tables for each country in the appendix}
      \item \st{AIC rows in the AIC table}
      \item \st{GDP comparison graph}
      \end{itemize}
\end{itemize}

@@

#################################################
# Calculates parameter estimates and confidence intervals
# for the CES production function given a country and an energyType.
#
# countryAbbrev is a string containing the 2-letter abbreviation for the country, e.g. "US" or "CN"
# energyType is a string, one of "Q", "X", or "U"
#
# returns a data.frame of data for the LINEX model.
# First row is the +95% CI on all parameters
# Second row contains the parameter estimates
# Third row is the -95% CI on all parameters
# Each column has names: gamma, lambda, delta_1, delta, rho_1, rho
##
cesData <- function(countryAbbrev, energyType){
  #First, check to see if we want useful work (U) AND one of the countries for which we don't have data.
  if (!haveDataCD(countryAbbrev, energyType)){
    #Return a column of NAs if the above conditions have been met.
    nRows <- 3 # +95% CI, CDe, and -95% CI.
    nCols <- 6 # gamma, lambda, delta_1, delta, sigma_1, and sigma
    df <- as.data.frame(matrix(NA, ncol = nCols, nrow = nRows))
    colnames(df) <- c("gamma", "lambda", "delta_1", "delta", "sigma_1", "sigma")
    rownames(df) <- c("+95% CI", "CES", "-95% CI")
    return(df)
  }
  # We have a combination of country and energy type for which we have data.
  modelCES <- cesModel(countryAbbrev, energyType)
  summaryCES <- summary(modelCES) # Gives a summary table.
#print(summaryCES)
  # Calculate the degrees of freedom as N_obs - N_params
  dofCES <- nrow(loadAndPrepDataForCES(countryAbbrev, energyType)) - 6 # Gives the degrees of freedom for the model.
#print(dofCES)
  tvalCES <- qt(ciHalfLevel, df = dofCES)
#print(tvalCES)
  # Calculate confidence intervals for the parameters of the model
  ciCES <- confint(modelCES, level = ciLevel)  # Calculates confidence intervals for the CES model.
#print(ciCES)
  # Get confidence intervals for the elasticities of substitution
  sigma_1.est <- deltaMethod(modelCES, "1.0 / (1.0 + rho_1)") # Estimates sigma_1 and its standard error (SE).
#print(sigma_1.est)
  sigma_1CI <- with(sigma_1.est, Estimate + c(-1.0, 1.0) * tvalCES * SE)
#print("sigma_1CI:")
#print(sigma_1CI)
  sigma.est <- deltaMethod(modelCES, "1.0 / (1.0 + rho)") # Estimates sigma and its standard error (SE)
#print(sigma.est)
  sigmaCI <- with(sigma.est, Estimate + c(-1.0, 1.0) * tvalCES * SE)
#print("sigmaCI:")
#print(sigmaCI)
  # Report results with SE
  gamma <- coef(modelCES)["gamma"] #as.numeric(coef(modelCES)["gamma"])
  lambda <- coef(modelCES)["lambda"] #as.numeric(coef(modelCES)["lambda"])
  delta_1 <- coef(modelCES)["delta_1"]
  delta <- coef(modelCES)["delta"]
  sigma_1 <- sigma_1.est$Estimate
  sigma <- sigma.est$Estimate
  # Combine all estimates and their confidence intervals into data frames with intelligent row names
  estCES <- data.frame(gamma=gamma, lambda=lambda, delta_1=delta_1, delta=delta, sigma_1=sigma_1, sigma=sigma)
  row.names(estCES) <- "CES"
#print(estCES)
  # The [1] subscripts pick off the lower confidence interval
#print(ciCES["gamma","2.5 %"])
  lowerCES <- data.frame(gamma=ciCES["gamma","2.5 %"], lambda=ciCES["lambda", "2.5 %"],
                         delta_1=ciCES["delta_1", "2.5 %"], delta=ciCES["delta", "2.5 %"],
                         sigma_1=sigma_1CI[1], sigma=sigmaCI[1])
  row.names(lowerCES) <- "-95% CI"
#print(lowerCES)
  # The [2] subscripts pick off the upper confidence interval
  upperCES <- data.frame(gamma=ciCES["gamma","97.5 %"], lambda=ciCES["lambda", "97.5 %"],
                         delta_1=ciCES["delta_1", "97.5 %"], delta=ciCES["delta", "97.5 %"],
                         sigma_1=sigma_1CI[2], sigma=sigmaCI[2])
  row.names(upperCES) <- "+95% CI"
#print(upperCES)
  # Now create the data for a table and return it
  dataCES <- rbind(upperCES, estCES, lowerCES)
#print("dataCES")
#print(dataCES)
  return(dataCES)
}

############
# Creates a row for the CES parameters table for the given country (2-letter code) and energyType (Q, X, or U)
##
cesCountryRow <- function(countryAbbrev, energyType){
  dataCES <- cesData(countryAbbrev, energyType)
  out <- cbind(dataCES["-95% CI", "gamma"], dataCES["CES", "gamma"], dataCES["+95% CI", "gamma"],
               dataCES["-95% CI", "lambda"], dataCES["CES", "lambda"], dataCES["+95% CI", "lambda"],
               dataCES["-95% CI", "delta_1"], dataCES["CES", "delta_1"], dataCES["+95% CI", "delta_1"],
               dataCES["-95% CI", "delta"], dataCES["CES", "delta"], dataCES["+95% CI", "delta"],
               dataCES["-95% CI", "sigma_1"], dataCES["CES", "sigma_1"], dataCES["+95% CI", "sigma_1"],
               dataCES["-95% CI", "sigma"], dataCES["CES", "sigma"], dataCES["+95% CI", "sigma"]
               )
  return(out)
}

###########################################
# Creates a number of rows in a data.frame that contain information 
# about the coefficients of a CES model for countryName and energyType
# Each parameter has its own row with confidence intervals.
# The country name is in a column. Which parameter is involved is
# also in a column.
# 
# The return type is a data.frame.
##
cesCountryRowsForParamsGraph <- function(countryName, energyType){
  #Create six rows, one for each parameter. Each row is a data.frame so that it is plottable!
  dataCES <- cesData(countryName, energyType)
  gammaRow <- data.frame(country = countryName, 
                          parameter = "gamma", 
                          lowerCI = dataCES["-95% CI", "gamma"], 
                          value = dataCES["CES", "gamma"], 
                          upperCI = dataCES["+95% CI", "gamma"])
  lambdaRow <- data.frame(country = countryName, 
                         parameter = "lambda",
                         lowerCI = dataCES["-95% CI", "lambda"],
                         value = dataCES["CES", "lambda"],  
                         upperCI = dataCES["+95% CI", "lambda"])
  delta_1Row <- data.frame(country = countryName, 
                         parameter = "delta_1",
                         lowerCI = dataCES["-95% CI", "delta_1"],
                         value = dataCES["CES", "delta_1"],  
                         upperCI = dataCES["+95% CI", "delta_1"])
  deltaRow <- data.frame(country = countryName, 
                         parameter = "delta",
                         lowerCI = dataCES["-95% CI", "delta"],
                         value = dataCES["CES", "delta"],  
                         upperCI = dataCES["+95% CI", "delta"])
  sigma_1Row <- data.frame(country = countryName, 
                         parameter = "sigma_1",
                         lowerCI = dataCES["-95% CI", "sigma_1"],
                         value = dataCES["CES", "sigma_1"],  
                         upperCI = dataCES["+95% CI", "sigma_1"])
  sigmaRow <- data.frame(country = countryName, 
                         parameter = "sigma",
                         lowerCI = dataCES["-95% CI", "sigma"],
                         value = dataCES["CES", "sigma"],  
                         upperCI = dataCES["+95% CI", "sigma"])
  table <- rbind(gammaRow, lambdaRow, delta_1Row, deltaRow, sigma_1Row, sigmaRow)
  return(table)
}

############################
# Aggregates the CES results into a big data table for the given energyType.
##
cesParamsTable <- function(energyType){
  #Do rbind on the results of creating a row in the table for every country abbreviation that we know.
  dataCES <- do.call("rbind", lapply(countryAbbrevs, cesCountryRow, energyType=energyType))
  colnames(dataCES) <- c(" ", "$\\gamma$",    " ", 
                         " ", "$\\lambda$",   " ",
                         " ", "$\\delta_1$",  " ",
                         " ", "$\\delta$",    " ",
                         " ", "$\\sigma_1$",  " ",
                         " ", "$\\sigma$",    " ")
  rownames(dataCES) <- countryAbbrevs
  tableCES <- xtable(dataCES, 
                    caption=paste("CES model (with $", tolower(energyType), "$) parameters for 1980-2011 (US, UK, JP), 1991-2010 (CN and ZA), and 1991-2011 (SA, IR, TZ, and ZM). (Parameter estimates beneath symbol. 95\\% confidence interval bounds to left and right.)", sep=""), 
                    label=paste("tab:CES_Parameters_With_", energyType, sep=""),
                    digits = c(1, 2,2,2, 2,2,2, 2,2,2, 2,2,2, 2,2,2, 2,2,2),
                    align = "r|ccc|ccc|ccc|ccc|ccc|ccc") #Sets alignment of the numbers in the columns
  return(tableCES)
}

############################
# Prints a table with parameters from a CES model for the given energyType. 
##
printCESParamsTable <- function(energyType){
  print(cesParamsTable(energyType), 
    caption.placement="top", 
    sanitize.colnames.function = identity, 
    size="\\tiny",
    table.placement="H")
}
@

@@

<<CES_Params_Table, results='asis', eval=TRUE>>=
printCESParamsTable("Q"); printCESParamsTable("X"); printCESParamsTable("U")
@
